---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
--- 
 

## Loading and preprocessing the data
###Import Data

```{r}
rawdata<-read.csv("activity/activity.csv")

#New data set with NA's removed
filterNA<-complete.cases(rawdata)
rawdataNoNA<-rawdata[filterNA,]
```


## What is mean total number of steps taken per day?
###Histogram of total number of steps taken each day (NA's removed)
I use the **sqldf** package for aggregating data!
```{r}
library("sqldf")
aggdata<-sqldf("SELECT date, sum(steps) AS totalsteps 
                FROM rawdataNoNA 
                GROUP BY date
                ORDER BY date")
hist(aggdata$totalsteps, main = "Histogram of Total Steps", xlab = "Total Steps Per Day")
````

###The mean and median of steps taken per day
```{r, echo= FALSE} 

paste("The mean is ",mean(aggdata$totalsteps),sep ="")

paste("The median is ",median(aggdata$totalsteps),sep="")

```


## What is the average daily activity pattern?
###Time Series Plot of steps averaged across days
First I calcuated the averages, then made the graph below using ggplot2.

```{r, echo=FALSE}
library(ggplot2)
aggdata2<-sqldf("SELECT interval, avg(steps) AS avgsteps 
                FROM rawdata 
                GROUP BY interval
                 ORDER BY interval ")

max<-sqldf("select interval, avgsteps
           FROM aggdata2
           ORDER BY avgsteps desc
           LIMIT 1")

#line graph
ggplot(aggdata2, aes(x = interval, y = avgsteps)) +geom_line() +geom_point(data = max)+geom_text(data = max, aes(label=avgsteps))

```

The interval that contains the max average steps is 835.


## Imputing missing values
The number of NA's is 

```{r, echo = FALSE}
dim(rawdata)[1]-dim(rawdataNoNA)[1]
```

I will use the mean for the 5-minute intervals to replace the NA's. Below is the code. The dataset cleanupNA will all the NA's replaced with the mean value of the corresponding interval

```{r}

#Use Mean's to replace NA's
cleanupNA<-merge(rawdata,aggdata2, by="interval")

i<-1
loopend<-dim(cleanupNA)[1]

for (i in 1:loopend) {
    if (is.na(cleanupNA[i,2])==TRUE){
        cleanupNA[i,2]<-cleanupNA[i,4]
    }
}


cleanupNA<-sqldf("SELECT * FROM cleanupNA ORDER BY date, interval, steps")

```

Below is the new histogram

```{r, echo = FALSE}
#new total steps per day
cleanupNAaggdata<-sqldf("SELECT date, sum(steps) AS totalsteps 
                FROM cleanupNA 
                GROUP BY date
               ORDER BY date")

#histogram of new data set
hist(cleanupNAaggdata$totalsteps, main = "Histogram of Total Steps", xlab = "Total Steps Per Day")


paste("The new mean is ",mean(cleanupNAaggdata$totalsteps),sep ="")

paste("The new median is ",median(cleanupNAaggdata$totalsteps),sep="")

```
After replacing the NA's with the mean of their respective interval, we see the median and the mean are equivalent now.

## Are there differences in activity patterns between weekdays and weekends?

###Weekdays vs. Weekends patterns
The factor variable has value **1** for **weekend** and **2** for **weekday**
```{r}
#add weekday factor
i <- 1
loopend<- dim(cleanupNA)[1]

#factor: 1 = weekend, 2 = weekday
for (i in 1:loopend){
    
    if (weekdays( as.Date(cleanupNA$date[i])) %in% c("Saturday","Sunday")){
        cleanupNA$factor[i] = "weekend"
    } else {
        cleanupNA$factor[i] = "weekday" 
    }
}

cleanupNA$factor <- as.factor(cleanupNA$factor)

```

Below is the graph of the average steps factored across weekdays or weekends

```{r, echo = FALSE}

cleanupNAaggdata2<-sqldf("SELECT factor, interval, avg(steps) AS avgsteps 
                FROM cleanupNA 
                GROUP BY factor,interval
                ORDER BY factor, interval")

#plot line graphs
ggplot(cleanupNAaggdata2, aes(x = interval, y = avgsteps)) +geom_line(aes(color = factor)) +facet_grid(factor~.)

```

Based on the plots, it looks like the number of stephs in the earlier part of the day are much higher for weekdays and as they day progresses, they taper off.
For weekends however, they number of steps are a bit lower initially, however they stay consistent throughout the majority of the day.
